---
title: "P8104 HW 1 LEL2176"
output: github_document
---

```{r, echo = FALSE}
#load in library and dataset
library(tidyverse)
library(moderndive)
data("early_january_weather")
```

## Problem 1
Write a short description of the dataset using inline R code; accessing the dataset help file can be informative. In your discussion, please include:
the variables in this dataset, including names / values of important variables
the size of the dataset (using nrow and ncol)
the mean temperature
Make a scatterplot of temp (y) vs time_hour (x); color points using the humid variable (adding color = ... inside of aes in your ggplot code should help). Describe patterns that are apparent in this plot.


the the `early_january_weather` dataset, the most important variables are `temp`,
`dewp`, `humid`, `wind_dir`, `wind_speed`, `wind_gust`, `precip`, `visib`, and 
`pressure`, which are all recorded according to `year`, `month`, `day`, and `hour`
and `time_hour`. we are able to observe fairly consistent values for all variable
with a few outliers. there are missing values for the `wind_gust` column, 
and all 0 values for the `precip` column. the temperatures seem to be in the upper 30s
and low 40s. the dewp is in the high 20s. the humid is in the upper 50s and low 60s. 
wind direction is in the mid 200s and wind speed varys from aroudn 8 to 15. the 
pressure stays fairly constant aroudn 1012. the rows of the data set
`nrow(early_january_data)`is 358. the columns of the data set 
`ncol(early_january_data)` is 15. the mean temperature of the data is: 
`mean(early_january_weather[["temp”]]` which is equal to 39.58212.

```{r}
weather_scat = ggplot(early_january_weather, aes(x = time_hour, y  = temp, color = humid)) + geom_point()
weather_scat
ggsave("weather_scat.pdf")
```

the scatter plot `weather_scat` shows the relationship between time measure by
hours and the recorded temperature. when analyzing the graph there is a clear
sinusoidal wave pattern that repeats everyday. these waves represents the highs and
lows of daily temperature with the temperature peaking midday and dropping during
the night when the sun is gone. however, the overall trend of the line has an 
increasing slope showing that the temperature is slowly increasing as the days go
on.


## Problem 2
This problem is intended to emphasize variable types and introduce coercion; some awareness of how R treats numeric, character, and factor variables is necessary for working with these data types in practice.

Create a data frame comprised of:

a random sample of size 10 from a standard Normal distribution
a logical vector indicating whether elements of the sample are greater than 0
a character vector of length 10
a factor vector of length 10, with 3 different factor “levels”
Try to take the mean of each variable in your dataframe. What works and what doesn’t?

In some cases, you can explicitly convert variables from one type to another. Write a code chunk that applies the as.numeric function to the logical, character, and factor variables (please show this chunk but not the output). What happens, and why? Does this help explain what happens when you try to take the mean?


```{r}
hw1_df = 
  tibble(
    x = rnorm(10),
    vec_log = c(x>0),
    vec_char = c("Child", "Adult", "Old Age", "Old Age", "Adult", "Old Age",
                 "Child", "Adult", "Adult", "Child"),
    vec_fact = factor(vec_char, levels = c("Child", "Adult", "Old Age"))
  )
```

When trying to take the mean of each variable of my data frame, I was able to obtain a numerical answer for `x` (my random sample) and `vec_log` (my logical vector), 
but not `vec_char` (my character vector) or `vec_fact` (my factor vector). 
I found my mean by doing `mean(hw1_df[[“variable_name”]])` but I also could have
implemented the `pull()` funciton in order to pull out specific variables from 
the dataframe. example: `pull(hw1_df, "vec_fact")` would give me just the factor
vector from the dataframe.

```{r}
num_log = as.numeric(hw1_df[["vec_log"]])
num_char = as.numeric(hw1_df[["vec_char"]])
num_fact = as.numeric(hw1_df[["vec_fact"]])
```

when using  the `as.numeric` function, I am able to output the vectors from my dataframe as numeric vectors. `as.num(hw1_df[["vec_log"]])` outputs my logical vector as a vector of numbers, which makes sense why I was able to calculate the mean for this vector. `as.numeric(hw1_df[["vec_char"]])` outputs a vector of only NA values because it was not able to convert the vector into numerical values, which makes sense why I was not able to calculate the mean of this vector. `as.numeric(hw1_df[["vec_fact"]])` outputs a vector with numerical values, which does not make sense because I am not able to calculate the mean of this vector, yet it is able to be converted into numerical values. however, I am able to find the mean of `vec_fact` using the newly created numerical vector: `mean(num_fact)` = 2.

