---
title: "p8105_hw5_lel2176"
author: "Lauren Lazaro"
date: "2023-12-01"
output: github_document
---

```{r}
library(tidyverse)
library(rnoaa)
library(janitor)
library(MASS)
library(modelr)
```


# Problem 1
The boostrap is helpful when you’d like to perform inference for a parameter / value / summary that doesn’t have an easy-to-write-down distribution in the usual repeated sampling framework. We’ll focus on a simple linear regression with tmax as the response with tmin and prcp as the predictors, and are interested in the distribution of two quantities estimated from these data:

r^2
log(β̂ 1∗β̂ 2)
Use 5000 bootstrap samples and, for each bootstrap sample, produce estimates of these two quantities. Plot the distribution of your estimates, and describe these in words. Using the 5000 bootstrap estimates, identify the 2.5% and 97.5% quantiles to provide a 95% confidence interval for r^2
 and log(β̂ 1∗β̂ 2)
Note: broom::glance() is helpful for extracting r^2
 from a fitted regression, and broom::tidy() (with some additional wrangling) should help in computing log(β̂ 1∗β̂ 2)

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2022-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
   dplyr::select(name, id, everything())

bootstraps = weather_df |> 
  modelr::bootstrap(n = 5000) |> 
  mutate(
    model = map(strap, \(df) lm(tmax ~ tmin + prcp, data = df)),
    result = map(model, broom::tidy),
    r2 = map(model, broom::glance)) |> 
  dplyr::select(result, r2) %>% unnest(r2) |> 
  dplyr::select(result, r.squared) %>% unnest(result) |> 
  dplyr::select(term, estimate, r.squared) 

bootstraps |> 
  filter(term == "(Intercept)") |> 
  ggplot(aes(x = r.squared)) +
  geom_density()

CI = bootstraps |> 
  filter(term == "(Intercept)") |> 
  dplyr::select(r.squared) |> 
  summarize(lower = quantile(r.squared, 0.025), 
            upper = quantile(r.squared, 0.975))

CI
```

The data is centered around 0.92 and is skewed to the left. We are 95% confident that the R-squared value is between `r CI$lower` and `r CI$upper`

```{r}

bootstraps |> 
  pivot_wider(names_from = term, values_from = estimate) |>  
  rename(b1 = tmin, 
         b2 = prcp) |>  
  mutate(logs = log(b1 * b2)) |>  
  ggplot(aes(x = logs)) + 
  geom_density()


CI_b1b2 = bootstraps |> 
  pivot_wider(names_from = term, values_from = estimate) |>  
  rename(b1 = tmin, 
         b2 = prcp) |>  
  mutate(logs = log(b1 * b2)) |> 
  summarize(NA_num = sum(is.na(logs)),
            lower = quantile(logs, 0.025, na.rm = T), 
            upper = quantile(logs, 0.975, na.rm = T))

CI_b1b2
```

The data skewed to the left with a peak in between -6 and -5.


## Problem 2

In this problem, you will analyze data gathered to understand the effects of several variables on a child’s birthweight. This dataset, available here, consists of roughly 4000 children and includes the following variables:

Propose a regression model for birthweight. This model may be based on a hypothesized structure for the factors that underly birthweight, on a data-driven model-building process, or a combination of the two. Describe your modeling process and show a plot of model residuals against fitted values – use add_predictions and add_residuals in making this plot.

Compare your model to two others:

One using length at birth and gestational age as predictors (main effects only)
One using head circumference, length, sex, and all interactions (including the three-way interaction) between these
Make this comparison in terms of the cross-validated prediction error; use crossv_mc and functions in purrr as appropriate.

Note that although we expect your model to be reasonable, model building itself is not a main idea of the course and we don’t necessarily expect your model to be “optimal”.

```{r}
birthweight =
  read_csv("./data/birthweight.csv") |> 
  mutate(
    babysex = as.factor(babysex),
    babysex = fct_recode(
      babysex, 
      "male" = "1", 
      "female" = "2"
      ),
    frace = as.factor(frace),
    frace = fct_recode(
      frace, 
      "white" = "1", 
      "black" = "2", 
      "asian" = "3", 
                       
      "puerto rican" = "4", 
      "other" = "8"),
    malform = as.logical(malform),
    mrace = as.factor(mrace),
    mrace = fct_recode(
      mrace, 
      "white" = "1", 
      "black" = "2", 
                       
      "asian" = "3", 
      "puerto rican" = "4")
    ) |> 
  drop_na()

lm =
  lm(bwt ~., data = birthweight)

step =
  stepAIC(lm, direction = "backward", trace = FALSE)

step |> 
  broom::tidy() |> 
  knitr::kable()

birthweight |>  
  add_predictions(step) |>  
  add_residuals(step) |>  
  ggplot(aes(x = pred, y = resid)) + 
  geom_point() +
  labs(title = "Fitted Values Vs. Residual Values", 
       x = "Fitted Values",
       y = "Residual Values")

comp1 =
  lm(bwt ~ blength + gaweeks, data = birthweight) |> 
  broom::tidy() |> 
  knitr::kable()

comp2 =
  lm(bwt ~ bhead*blength*babysex, data = birthweight) |> 
  broom::tidy() |> 
  knitr::kable()

CV =
  crossv_mc(birthweight, 100) |> 
  mutate(
    test = map(test, as_tibble),
    train = map(train, as_tibble) 
    )

comp_CV =
  CV |>  
  mutate(
    mod1  = map(
      train, 
      ~lm(bwt ~ babysex + bhead + blength + delwt + fincome + 
    gaweeks + mheight + mrace + parity + ppwt + smoken, data = .x)),
    mod2  = map(train, ~lm(bwt ~ gaweeks + blength, data = .x)),
    mod3  = map(train, ~lm(bwt ~ bhead*blength*babysex, data = .x))
    ) |>  
  mutate(
    rmse1 = map2_dbl(mod1, test, ~rmse(model = .x, data = .y)),
    rmse2  = map2_dbl(mod2, test, ~rmse(model = .x, data = .y)),
    rmse3 = map2_dbl(mod3, test, ~rmse(model = .x, data = .y))
    ) |>  
  summarize(
    avg_mod1 = mean(rmse1),
    avg_mod2 = mean(rmse2),
    avg_mod3 = mean(rmse3)
    )

comp_CV

CV |>  
  mutate(
    mod1  = 
      map(train, ~lm(bwt ~ babysex + bhead + blength + delwt + fincome + 
                       gaweeks + mheight + mrace + parity + ppwt + smoken, 
                     data = .x)),
         mc1  = map(train, ~lm(bwt ~ gaweeks + blength, data = .x)),
         mc2  = map(train, ~lm(bwt ~ bhead*blength*babysex, data = .x))
    ) |>  
  mutate(
    rmse1 = map2_dbl(mod1, test, ~rmse(model = .x, data = .y)),
         rc1  = map2_dbl(mc1, test, ~rmse(model = .x, data = .y)),
         rc2 = map2_dbl(mc2, test, ~rmse(model = .x, data = .y))
    ) |> 
  dplyr::select(rmse1, rc1, rc2) |>   
  pivot_longer(everything(),
               names_to = "model", 
               values_to = "rmse",
               names_prefix = "rmse_") |> 
  mutate(
    model = fct_inorder(model)
    ) |>  
  ggplot(aes(x = model, y = rmse)) + 
  geom_violin(aes(fill = model))
```
