---
title: "p8105_hw3_lel2176"
output: github_document
date: "2023-10-13"
---

```{r}
library(tidyverse)
library(p8105.datasets)
library(dplyr)
data("brfss_smart2010")
```

## Problem 2

This problem uses the BRFSS data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets package.

First, do some data cleaning:

format the data to use appropriate variable names;
focus on the “Overall Health” topic
include only responses from “Excellent” to “Poor”
organize responses as a factor taking levels ordered from “Poor” to “Excellent”
Using this dataset, do or answer the following (commenting on the results of each):

In 2002, which states were observed at 7 or more locations? What about in 2010?
Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).
Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r}
BRFSS = 
brfss_smart2010 |> 
  select(Year, Locationabbr, Locationdesc, Class, Topic, Question, Response, Sample_Size, Data_value, Confidence_limit_Low, Confidence_limit_High, Display_order, GeoLocation) |> 
  rename(year = Year) |> 
  rename(state = Locationabbr) |> 
  rename(county = Locationdesc) |> 
  rename(class = Class) |> 
  rename(topic = Topic) |> 
  rename(question = Question) |> 
  rename(response = Response) |> 
  rename(sample_size = Sample_Size) |> 
  rename(data_value = Data_value) |> 
  rename(lower_confidence_lim = Confidence_limit_Low) |> 
  rename(upper_confidence_lim = Confidence_limit_High) |> 
  rename(display_order = Display_order) |> 
  rename(geolocation = GeoLocation)
```

```{r}
overall_health = BRFSS[BRFSS$topic=="Overall Health",]

sorted_overall_health = 
overall_health |> 
  mutate(overall_health = fct_relevel(overall_health$response, c("Poor", "Fair", "Good", "Very good", "Excellent"))) |> 
  arrange(overall_health)

sorted_overall_health
```

```{r}
counties_2002 = 
  sorted_overall_health |> 
  filter(year == 2002) |> 
  group_by(state) |> 
  summarize(counties = length(unique(county))) |> 
  filter(counties >= 7)

counties_2002
```

```{r}
counties_2010 = 
  sorted_overall_health |> 
  filter(year == 2010) |> 
  group_by(state) |> 
  summarize(counties = length(unique(county))) |> 
  filter(counties >= 7)

counties_2010
```

States with 7 or more observed `Excellent` responses 2002:
CT, FL, MA, NJ, NC, PA

States with 7 or more observed `Excellent` responses 2010:
CA, CO, FL, MD, MA, NE, NJ, NY, NC, OH, PA, SC, TX, WA

```{r}
excellent_data = 
  sorted_overall_health |> 
  filter(response == "Excellent") |> 
  group_by(year, county) |> 
  summarize(data_value_average = mean(data_value, na.rm = TRUE))

excellent_data

ggplot(
  excellent_data, aes(x = year, y = data_value_average, group = county))+
  geom_point() +
  geom_line() +
  labs(
    title = "'Excellent' Responses by State Over Time", 
    x = "year", 
    y = "average"
  )
```

this spaghetti plot shows the average number of `Excellent` responses to a question about overall health throughout the years by each state. most of the states follow similar trend lines with a few high and low outliers. there appears to be an ever so slight decrease in `Excellent` responses from 2002 to 2010. 

```{r}
data_2006_2010 =
  sorted_overall_health |> 
  filter(state == "NY" & (year == "2006" | year == "2010")) |> 
  ggplot(aes(x = data_value)) +
  geom_histogram(binwidth = 10)  +
  facet_grid(year ~ response)

data_2006_2010
```

this visualization of data shows the distribution of `data_value` for the different responses for all of the different locations in New York in both 2006 and 2010. it appears that in 2010, there are more observed `Excellent`, `Fair`, `Good`, and `Poor` responses than in 2006. the observations for `Very good` are about the same for 2006 and 2010. 

## Probem 3

Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices can measure MIMS in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses accelerometer data collected on 250 participants in the NHANES study. The participants’ demographic data can be downloaded here, and their accelerometer data can be downloaded here. Variables *MIMS are the MIMS values for each minute of a 24-hour day starting at midnight.

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
demographics = 
  read_csv("data/nhanes_covar.csv")[-c(1, 2, 3, 4),] 
demographics = na.omit(demographics)

demographics = 
  demographics |> 
  rename(seqn = `...1`) |> 
  rename(sex = `1 = male`) |> 
  rename(age = `...3`) |> 
  rename(bmi = `...4`) |> 
  rename(education = `1 = Less than high school`) |> 
  mutate(
    sex = case_match(
      sex, 
      "1" ~ "male",
      "2" ~ "female"
    )
  ) |> 
  mutate(
    education = case_match(
      education, 
      "1" ~ "less than high school",
      "2" ~ "high school equivalent",
      "3" ~ "more than high school"
    )
  ) |> 
  mutate(seqn = as.numeric(seqn)) |> 
  mutate(sex = as.factor(sex)) |> 
  mutate(age = as.numeric(age)) |>  
  mutate(bmi = as.numeric(bmi)) |>  
  mutate(education = as.factor(education))

demographics = demographics[demographics$age > 21,]


accelerometer = read_csv("data/nhanes_accel.csv")

accelerometer |> 
  janitor::clean_names() 

accelerometer = 
  na.omit(accelerometer)
```

```{r}
demo_acc =
  demographics |> 
  group_by(education, sex) |> 
  summarize(count = n()) |> 
  pivot_wider(names_from = sex, values_from = count)

  knitr::kable(demo_acc)
```

```{r}
age_plot =
  demographics |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 7) +
  facet_grid(sex ~ education) +
  labs(
    title = "Distribution of Age According to Sex and Education",
    x = "Age",
    y = "Count"
  )

print(age_plot)
```

the table `demo_acc` shows the male and female distribution of different education levels. the different education levels include `high school equivalent`, `less than high school`, and `more than high school`. from this table, we see more males with `high schoool equivalent` education, but more females for both `less than high school` and `more than high school` educations. the plot `age_plot` shows a visual representation of the male and female distribution of different education levels. through this graph, we can compare each age group of males and females for the three various education levels. `age_plot` follows the same trends as seen in `demo_acc`, but more detail is shown since age is represented on the x-axis. 

```{r}
acc_wo_seqn = accelerometer[,-1] 

total_activity_df = 
  accelerometer |> 
  rename(seqn = SEQN) |> 
  mutate(total_activity = rowSums(acc_wo_seqn))

total_activity_df

join_data =
  demographics |> 
  inner_join(total_activity_df, by = "seqn")

join_data

ggplot(join_data, aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  facet_grid(. ~ education) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Trends in Activity Level According to Gender and Education",
    x = "age",
    y = "total activity"
  )

```

the graph showing `trends in activity level according to gender and education` consistently shows in each education level group that activity level decreases as age increases. overall, as individuals age, their activity goes down. however, the slope at which `total_activity` decreases is the steepest for the `less than high school` education level and is the least steep for the `more than high school` education level. the general trends are mostly similar for both males and females in each education level group. 

```{r}
join_data = 
  join_data |>  
  pivot_longer(cols = starts_with("min"), names_to = "min", values_to = "activity") |>  
  mutate(min = str_remove(min, "min")) |>  
  mutate(min = as.integer(min)) |>  
  ggplot(aes(x = min, y = activity, color = sex)) +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_wrap(~education)
  
join_data

```

the plot `join_data` shows the distribution of male and female activity levels throughout the three various education levels more in depth using 24-hour activity time courses. the plot shows in each education level that the activity level rapidly increases at the start of the day, stays constant in the middle, and then decreases as time approaches the end of the 24 hours. in the `more than high school` education level group, there are many more outliers where individuals were significantly more active. these outliers are especially frequent among the males in this category. 
